### 2018-01-09 submission/manuscript.Rmd
change intro final paragrpah

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor for both adenoma and carcinoma. To accomplish this we first assessed whether diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. We then assessed how common CRC-associated genera (*Fusobacterium*, *Parvimonas*, *Peptostreptococcus*, or *Porphyromonas*) affect the RR of adenoma or carcinoma. Next, using Random Forest models, we analyzed whether the full community or only the CRC-associated genera resulted in better model classification based on the area under the curve (AUC). Our results suggest that the community changes as disease severity worsens and that this community is important for disease classification. However, since the changes in community were subtle for adenoma we also examined what effect and sample size the studies that were used were adequately powered for. Although we analyzed data sets that sampled large numbers of individuals, our results indicate the individual studies were underpowered for detecting effect size differences of 10% or below between the case and control groups.

>>>>

Using both feces (n = `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T)`) and tissue (`r sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` samples from `r length(unique(flemer_data$Individual)) + sum(filter(tissue_n, study == "lu")[, c("control", "adenoma")])/2 + sum(filter(tissue_n, study == "burns")[, c("control", "cancer")])/2 + sum(filter(tissue_n, study == "chen")[, c("control", "cancer")])/2 + round(sum(filter(tissue_n, study == "dejea")[, c("control", "cancer")])/2) + 1 + sum(filter(tissue_n, study == "geng")[, c("control", "cancer")])/2 +  sum(filter(tissue_n, study == "sana")[, c("control", "cancer")])` individuals) totaling over `r sum(stool_n$control) + sum(stool_n$adenoma, na.rm = T) + sum(stool_n$cancer, na.rm = T) + sum(tissue_n$control) + sum(tissue_n$adenoma, na.rm = T) + sum(tissue_n$cancer, na.rm = T)` total samples across 14 studies [@hale_shifts_2017; @flemer_tumour-associated_2017; @chen_human_2012; @wang_structural_2012; @lu_mucosal_2016; @brim_microbiome_2013; @weir_stool_2013; @dejea_microbiota_2014; @sanapareddy_increased_2012; @ahn_human_2013; @burns_virulence_2015; @zeller_potential_2014; @baxter_microbiota-based_2016; @geng_diversified_2013] [Table 1 & 2], we expand both the breadth and scope of the previous meta-analysis to investigate whether the bacterial community is an important risk factor or if select members can recapitulate the same information for both adenoma and carcinoma. To accomplish this we first assessed whether bacterial diversity changes throughout disease (control to adenoma to carcinoma) and if it results in an increased relative risk (RR) for adenoma or carcinoma. We then assessed what genera, if any, increase or decrease the RR of adenoma or carcinoma. Next, using Random Forest models, we analyzed whether the full community or only the top 5 increased and top 5 decreased RR genera combined resulted in better model classification, based on the area under the curve (AUC). Our results suggest that the community changes as disease severity worsens and that this community is important for disease classification. Finally, we also examined at what effect and sample size the studies that were used were powered for and the sample size needed to get to an 80% power. Our power analysis suggests that each individual study used was underpowered for detecting effect size differences of 10% or below between the case and control groups.

### 2018-01-09 submission/manuscript.tex
change intro final paragrpah

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor for both adenoma and carcinoma. To
accomplish this we first assessed whether diversity changes throughout
disease (control to adenoma to carcinoma) and if it results in an
increased relative risk (RR) for adenoma or carcinoma. We then assessed
how common CRC-associated genera (\emph{Fusobacterium},
\emph{Parvimonas}, \emph{Peptostreptococcus}, or \emph{Porphyromonas})
affect the RR of adenoma or carcinoma. Next, using Random Forest models,
we analyzed whether the full community or only the CRC-associated genera
resulted in better model classification based on the area under the
curve (AUC). Our results suggest that the community changes as disease
severity worsens and that this community is important for disease
classification. However, since the changes in community were subtle for
adenoma we also examined what effect and sample size the studies that
were used were adequately powered for. Although we analyzed data sets
that sampled large numbers of individuals, our results indicate the
individual studies were underpowered for detecting effect size
differences of 10\% or below between the case and control groups.

>>>>

Using both feces (n = 1737) and tissue (492 samples from 350
individuals) totaling over 2229 total samples across 14 studies
{[}12,16--28{]} {[}Table 1 \& 2{]}, we expand both the breadth and scope
of the previous meta-analysis to investigate whether the bacterial
community is an important risk factor or if select members can
recapitulate the same information for both adenoma and carcinoma. To
accomplish this we first assessed whether bacterial diversity changes
throughout disease (control to adenoma to carcinoma) and if it results
in an increased relative risk (RR) for adenoma or carcinoma. We then
assessed what genera, if any, increase or decrease the RR of adenoma or
carcinoma. Next, using Random Forest models, we analyzed whether the
full community or only the top 5 increased and top 5 decreased RR genera
combined resulted in better model classification, based on the area
under the curve (AUC). Our results suggest that the community changes as
disease severity worsens and that this community is important for
disease classification. Finally, we also examined at what effect and
sample size the studies that were used were powered for and the sample
size needed to get to an 80\% power. Our power analysis suggests that
each individual study used was underpowered for detecting effect size
differences of 10\% or below between the case and control groups.
